\chapter{Detailed industry and sector mappings}\label{app:industry-crosswalks}

The industries defined in the NED and AA modules were built from data classified using different systems. Some of the equivalencies are shown in the tables in Chapter \ref{sec:system-overview}. However, definitive listings span several pages in some instances, and collected in the Appendix:
\begin{itemize}
\item The NED industries correspond to several different industries and market segments, as shown in Table \ref{tab:ned-industry-definitions}. Note that the mappings are not unique in some instances, as discussed below.
\item Four-digit NAICS codes or ranges thereof unambiguously map to a single NED industry, with the exception of construction (which maps to four NED industries) and textile furnishings (NAICS 3141), which is split between two NED industries. This mapping is shown in Table \ref{tab:naics-to-ned}.
\item Each IMPLAN (IMpact for PLANning) sector unambiguously maps to a single NED industry, as shown in Table \ref{tab:implan-mappings}.
\item Each four-digit PUMS sectors maps to one or more NED industries, with some associated with three or more of them. These mappings are shown in Table \ref{tab:pums-to-ned}.
\end{itemize}

NAICS codes can be defined with two to six digits of detail. These levels are used interchangably within the definition of NED industries. For example, construction is defined in terms of the two-digit code ``23'' (i.e., encompassing all of the more detailed three to six-digit codes beneath them). Electrical energy generation, on the other hand, only includes a single four-digit code.

The PUMS sectors are only defined as four-digit codes. However, with few exceptions each PUMS sector maps to more than one NED industry. Thus, only the mapping from PUMS sectors is included. 

\newpage
\input{appendices/ned-industry-definitions.tex}
\input{appendices/naics-to-ned.tex}
\input{appendices/implan-mappings.tex}
\input{appendices/pums-to-ned.tex}
